---
title: "Math 472/572 Homework 5"
author: "Graham Swain"
date: "3/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Functions in R
There are 40 possible points on this homework.  It is **due by Friday, April 8, 
2022 at 11:59 p.m.**  You may talk to your classmates about the homework, but
what you submit MUST be an indication of your own work and understanding.
Please contact Dr. McNelis if you have any questions.

*Problems*:

1.  [5 points] Create a function in R called `scale_ab()` that takes two inputs, 
    a numerical vector, x, and a vector that contains the values of a 
    and b (where a is assumed to be the first element and b the second element).
    Your function should output a rescaled version of x such that its values 
    have been rescaled to be between a and b, so the smallest value maps to a 
    and the largest to b.  This can be accomplished using the formula
    \begin{equation*}
    x_{new} = a + \frac{x_{old} - \alpha}{\beta - \alpha} (b - a)
    \end{equation*}
    
    where $\alpha$ is the minimum value of x and $\beta$ is the maximum value
    of x.  NOTE: Your function should have a default argument that na.rm is
    TRUE, and should pass the value stored in this na.rm to any necessary
    functions as it's na.rm input.
    
```{r 1}
scale_ab = function ( x, ab, na.rm = TRUE ) {
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  
  scaled <- vector(typeof(x), length(x))
  
  min <- min(x)
  max <- max(x)
  
  (scaled <- ab[1] + ((x - min) / (max - min)) * (ab[2] - ab[1]))
}
```
    
2.  [5 points] Use a single string of commands to map your `apply_ab` to all 
    numerical variables in the `iris` data frame, with a = -1 and b = 1, and 
    the use the `pairs` function to make a scatter plot of all possible pairs 
    of numerical variables.

```{r 2}
sapply(iris[, sapply(iris, is.numeric)], scale_ab, ab = c(-1, 1)) %>% pairs
```
    
3.  [5 points]  Use the following code to randomly generate a tibble of fake 
    test scores, then make a `pivot_longer` version of the tibble to help you 
    create an illustration of the distribution of these test grades.  The 
    illustration should include histograms (using bin widths of 10) with a 
    density plot on top of them.  They should be broken up by test, and their 
    horizontal axes scales should all go from 0 to 100.  It is appropriate to 
    add a title to this illustration as well.  Dr. McNelis has provided an 
    example of such an image with this assignment.
    
```{r 3}
testscores <- tibble(
  test1 = sample(rnorm(100,78,15), 50),
  test2 = sample(c(rnorm(100,65,21),rep(NA,5)), 50),
  test3 = sample(c(rnorm(100,75,12),rep(NA,10)),50),
  test4 = sample(c(rnorm(100,82,9),rep(NA,10)),50)
)

fixed <- as.tibble(sapply(na.omit(testscores), scale_ab, ab = c(0, 100)))
testscores_longer <- (fixed %>%
                        pivot_longer(test1:test4, names_to = "Test", values_to = "Score") %>% 
                        na.omit())

ggplot(testscores_longer) + 
  geom_histogram(aes(x = Score, y = ..density.., fill = Test, color = Test), binwidth = 10, alpha = .5) +
  geom_density(aes(x = Score, color = Test)) +
  facet_wrap(~ Test) +
  labs(title = "Distribution of Test Scores")
```

    
4.  [5 points]  Make a function, `zscore`, that will take as input a vector of 
    numerical values, x, and an optional na.rm argument (with the default na.rm 
    set to be TRUE).  As with the `scale_ab` function in part 1, you should pass 
    the value stored in the na.rm input on to other appropriate values called 
    in the body of your function. This function is to return a vector of 
    corresponding $z$-score values.  Then apply this function to the test
    scores in `testscores` using either a for loop or a mapping.  Do not
    overwrite the `testscores` values.
    
```{r 4}
zscore <- function( x, na.rm = TRUE) {
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  z <- vector(typeof(x), length(x))
  mean <- mean(x, na.rm = na.rm)
  sd <- sd(x, na.rm = na.rm)
  
  (z <- (x - mean) / sd)
}

sapply(testscores, zscore)
```
    
    
5.  "Curving Grades":  At times faculty mention "curving" grades and that 
    likely means just adding a fixed amount of points to test grades.  But we
    will consider other ways to "curve" grades, which includes trying to 
    help them fit a more normal distribution.  The following subproblems will
    require you to create functions for each different form of curving that
    Dr. McNelis could think of.  Each of the functions you create, **with the
    exception of `curves_sqrt`**, should have
    an optional input of na.rm with the default value set to TRUE.  Treatment
    of NA values should be determined by the value of na.rm passed to the
    function.  You should make use of functions you've created already whenever
    appropriate.
    
    
    a.  [2 points]  Create a function called `curve_sqrt` that takes an input 
      of a numerical vector, x, and outputs a vector of nuerical values 
      calculated by
$$
        \begin{equation*}
        x_{new} = 10*\sqrt{x}
        \end{equation*}
$$
    
```{r 5a}
curve_sqrt <- function(x, na.rm = TRUE) {
  if (na.rm) {
    x = x[!is.na(x)]
  }
  
  new <- vector(typeof(x), length(x))
  (new <- 10 * sqrt(x))
}
```
    
      
    b.  [4 points]  Create a function called `curve_ms` that takes inputs of a 
      numerical vector, x, a desired mean for the distribution, mu, and a 
      desired standard deviation for the distribution, sigma.  This function 
      should output a vector of numerical values calculated by:
        i)  Calculating the z score of your x value, z_x, then
        ii) Convert it a distribution with the desired mean and standard deviation
            by calculating 
$$
            \begin{equation*}
            x_{new} = \mu + z_x * \sigma
            \end{equation*}
$$
            corresponding to adding the bump value to the each
            
```{r 5b}
curve_ms <- function(x, mu, sd, na.rm = TRUE) {
  if (na.rm) {
    x = x[!is.na(x)]
  }
  z_x = zscore(x, na.rm)
  new <- vector(typeof(x), length(x))
  
  (new <- mu + z_x * sd)
}
```

            
    c.  [4 points]  Create a function called `curve_ab` that takes inputs of a 
      numerical vector, x, the minimum curved score, and the maximum curved
      score. This function should output a vector of numerical values 
      calculated by:
        i)   Calculating the z score of your x value, z_x, then
        ii)  Finding the associated probability of getting a score no larger
             than that on the standard normal curve (i.e., find pnorm(z_x)), then
        iii) Send scale those values so they are distributed in the same manner
             over the interval [a,b]
             
```{r 5c}
curve_ab <- function(x, min, max, na.rm = TRUE) {
  if (na.rm) {
    x = x[!is.na(x)]
  }
  z_x = zscore(x, na.rm)
  p_x = pnorm(z_x)
  new <- vector(typeof(x), length(x))
  
  (new <- scale_ab(p_x, c(min, max)))
}
```
             
    d.  [4 points]  Create a function called `curve_linear` that takes inputs of 
      a numerical vector, x, the desired mean (mu_d), and the desired maximum 
      (M_d) for the curved test scores. This function should output a vector of 
      numerical values calculated by:
$$
        \begin{equation*}
        x_{new} = mu_d + \frac{(M_d - mu_d)}{(M_a - mu_a)}(x - mu_a)
        \end{equation*}
$$
        where $M_a$ = maximum value from x and $mu_a$ = mean of values in x.
        
```{r 5d}
curve_linear <- function(x, mu_d, M_d, na.rm = TRUE) {
  if (na.rm) {
    x = x[!is.na(x)]
  }
  mu_a = mean(x)
  M_a = max(x)
  new <- vector(typeof(x), length(x))
  
  (new <- mu_d + ((M_d - mu_d) / (M_a - mu_a)) * (x - mu_a))
}
```


6.  [6 points]  Apply your curving functions from Problem 5 to the following 
    tibble, saving each output in a new variable for the tibble with an 
    appropriate name (like "sqrt_curve", "new_mu_sigma", etc).  Specifics for 
    additional inputs are given below.  Once you have your new tibble, use what 
    you learned in Problem 3 to generate a plot of the histograms with density 
    plots for each version of the test scores.
      For `curve_ms` we want a mean of 75 and standard deviation of 8
      For `curve_ab` we want a min value to be 50 and a max values to be 100
      For `curve_linear` use a mean of 75 and a max of 100.

```{r}
test_grades <- tibble(
  original = c(rep(NA,2), 12, 38, rep(52,4), 53, 54, rep(55,6), 56, 58, 
               59, 59, 61, rep(62,5), 89)
)

sqrt_curve <- curve_sqrt(test_grades)
ms_curve <- curve_ms(test_grades, 75, 8)
ab_curve <- curve_ab(test_grades, 50, 100)
linear_curve <- curve_linear(test_grades, 75, 100)

curves <- tibble(sqrt_curve, ms_curve, ab_curve, linear_curve)
names(curves) <- c("sqrt", "ms", "ab", "linear")

curves_longer <- (curves %>%
                    pivot_longer(sqrt:linear, names_to = "Curve", values_to = "Grade"))

ggplot(curves_longer) + 
  geom_histogram(aes(x = Grade, y = ..density.., fill = Curve, color = Curve), binwidth = 10, alpha = .5) +
  geom_density(aes(x = Grade, color = Curve)) +
  facet_wrap(~ Curve) +
  labs(title = "Test Scores with Different Curves")
```

      