---
title: "Smooth"
author: "Graham Swain"
date: "1/11/2022"
output: html_document
---

## Loading Tidyverse Library
```{r setup, message=FALSE}
library(tidyverse)
```

## Looking at geom_smooth in ggplot
Section 3.6 of our online textbook 
<https://r4ds.had.co.nz/data-visualisation.html#geometric-objects>.
Previously we played with `ggplot(data = mpg) + geom_point()`.

```{r}
attach(mpg)
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Let's try adding a scatter plot to this graph. But those come from `geom_point`.
Have multiple geoms in one `ggplot`.

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

Let's change method from default of "loesss" to "lm".
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy), method = "lm") +
  geom_point(mapping = aes(x = displ, y = hwy))
```

With smooth, we can adjust line styles, which wasn't appropriate for our point option.
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) +
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))
```