---
title: "2022-02-15 Math 472.572 Wrangling Part 1"
author: "Erin McNelis"
date: "2/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#  Some New Discussion About Tibbles

Dr. McNelis has often referred to our data stored in tibbles as data
frames (`data.frame`).  The data frame is the earlier object used for storing
tables of data.  A tibble is a data frame, but it has been tweaked some to
improver performance and use.

We will typically import data from other sources (text files, spreadsheets, web pages, etc.) and give them a name in R.  The importing can import the
data as a tibble or as a data frame.  We'll focus on the `readr` library
later, which will typically output data as tibbles, but we can specify
tibbles (or data frames) by typing these in at a command line or in a
script file.

Consider putting the following in a data file and in a tibble, separately.
Let's suppose we want to make a set of three ficticious characters and list
their initials, their date of birth, their number of siblings, and their
high school GPA.  Consider randomly generating these values, such as using

`ints<-paste(sample(letters[1:24],4,replace=TRUE),sample(letters[1:24],4,replace=TRUE),sample(letters[1:24],4,replace=TRUE),sep="" )`
and
`bdays <- sample(seq(as.Date('1999/01/01'), as.Date('2015/01/01'), by="day"), 4)`
Think about what these may mean.

```{r}
inits <- paste(sample(letters[1:24], 4, replace= TRUE),
               sample(letters[1:24], 4, replace= TRUE),
               sample(letters[1:24], 4, replace= TRUE), 
               sep = "")
inits

bdays <- sample(seq(as.Date('1999/01/01'), as.Date('2015/01/01'), by = "day"), 4)
bdays

sibs <- sample(0:9, 4, replace = TRUE)
sibs

gpas <- sample(seq(0, 5, .01), 4, replace = TRUE)
gpas

gpasn <- sample(rnorm(4, mean = 3.14, sd = 1), replace = TRUE)
gpasn

df <- data.frame(inits = inits, bdays = bdays, sibs = sibs, gpas = gpas)
df

tb <- tibble(inits = inits, bdays = bdays, sibs = sibs, gpas = gpas)
tb
```

Investigate these by typing their names, and/or using the `str()` function.

```{r}
str(df)
str(tb)
head(df, 2)
head(tb, 2)
print(df, n = 2)
print(tb, n = 2)
```

## Viewing Values in a Tibble

We can see a column of value, i.e. a variable's values using $ or [] or [[]].

```{r}
# View the GPA's of each: df, tb

df$gpas
tb$gpas
tb[["inits"]] # specify name of variable
tb[[1]] # specify column of variable
tb[1] # outps as a tibble rather than a vector
tb[,1] # getting the first column
tb[1,] # getting the first row
```


## Converting Data Frames to Tibbles

```{r}
new_tb <- as_tibble(df)
new_tb
```


## Referring to Specific Values in a Tibble (by Variable or Entry)


### Using Pipes

```{r}
tb2 <- tibble(x = 2:6, y = sin(x), z = 32)
tb2

tb2 %>% .$y
tb2 %>% .[["x"]]
```

# Parsing

Sometimes the way your data is written may not be the way you want it...

```{r}
salary <- "14.389,72" # some countries use "." as a separator/grouping and "," as decimal indicator

# Want to read this as a decimal number
parse_number(salary, locale = locale(decimal_mark = ","))

# parse_number ignores all characters
fake_sal <- parse_number("I make $27,846,319.52 each year.")
fake_sal

fake_sal%%1

# Note, to see the value of fake_sal without potential limits of displaying values to the command line,
# check out the Environment tab to read the value stored in fake_sal

# Change encoding read
sentence <- "El Ni\xf1o is a weather pattern."
parse_character(sentence, locale = locale(encoding = "Latin1"))

parse_double(c("245,67", "-32,675", "182,918"), locale = locale(decimal_mark = ","))
```

```{r}
library(readr)
challenge <- read_csv("~/math472/class_week6/challenge.csv")
View(challenge)
tail(challenge, 4)
```


```{r}
weather <- read_excel("~/math472/class_week6/WeatherData.xlsx")

new_weather <- weather %>% mutate(temp = parse_number(Temperature), 
                   dew = parse_number(weather[["Dew Point"]]))
new_weather
```

# Web Scraping from Weather Service

```{r}
library(rvest)

forecast <- read_html("https://forecast.weather.gov/MapClick.php?lat=35.372700000000066&lon=-83.22461999999996")
forecast
View(forecast)

# pick off the nodes that have "temp"

seven_day <- forecast %>% html_nodes(".temp") %>% html_text()
seven_day
```

# Scraping Tables from Web Pages

```{r}
movies <- read_html("https://en.wikipedia.org/wiki/List_of_highest-grossing_films_in_the_United_States_and_Canada")
movies

tables <- movies %>% html_table(fill = TRUE)
# don't display it
# tables
first_table <- tables[[1]]
first_table

table1 <- first_table[-1,-1]
table1

library(janitor)
table1 <- table1 %>% clean_names()
table1

names(first_table)
names(table1)
View(table1)
str(table1) # money values are strings

table1 %>% 
  mutate(lifetime_gross = parse_number(lifetime_gross)) %>% 
  arrange(desc(lifetime_gross)) %>% 
  head(20)

table1 %>% 
  mutate(lifetime_gross = parse_number(lifetime_gross)) %>%
  arrange(desc(lifetime_gross)) %>% 
  head(20) %>%
  mutate(title = fct_reorder(title, lifetime_gross)) %>%
  ggplot() + 
  geom_bar(aes(x = lifetime_gross, y = title), stat = "identity", fill = "blue") +
  labs(title = "Top 20 Grossing Movies in US and Canada", caption = "Data Source: Wikipedia")
```


